Garbage Collection Evaluation Infrastructure for the Cog VM

### 2018-06-12 paper.tex
added sec 2-3

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

%%%
% Legend definition (so I can change it once for all)
%%%

\def\legBase{Baseline\xspace}
\def\legSmartSyntaxPlug{Syntax\xspace}
\def\legSlangPlug{Plugin\xspace}
\def\legSlang{Slang\xspace}
\def\legRTL{Slang+RTL\xspace}
\def\LLVMMac{LLVM-Mac\xspace}
\def\GCCLinux{GCC-Linux\xspace}
\def\plotHeight{6.8cm}

%%%
% End Legend
%%%

@@

>>>>

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

@@

Trying to build a more advanced algo, need to measure..
ref impl to compare
Try to build benchmarks 
->

\section{Reference Implementations}
\label{sec:ref}

To evaluate new full GC algorithms, we need to compare it against existing algorithms. Two of the most common GC algorithms are Mark-Compact and Mark-Sweep. We describe here the current Mark-Compact implementation, which has been in production for the past few years, and an implementation of Mark-Sweep we introduced as a reference earlier this year.

\subsection{Mark-Compact}

Cog's Mark-Compact algorithm, named \emph{SpurPlanningCompactor}, implements the classic planning compaction algorithm for Spur.  It uses the fact that there is room for a forwarding pointer in all objects to store the eventual position of an object in the first field. It therefore first locates a large free chunk, or eden or a memory segment, to use as the savedFirstFieldsSpace, which it uses to store the first fields of objects that will be compacted. It then makes at least three passes through the heap.

The first pass plans where live movable objects will go, copying their first field to the next slot in savedFirstFieldsSpace, and setting their forwarding pointer to point to their eventual location (see planCompactSavingForwarders). The second pass updates all pointers in live pointer objects to point to objects' final destinations, including the fields in savedFirstFieldsSpace (see updatePointers and updatePointersInMobileObjects). The third pass moves objects to their final positions, unmarking objects, and restoring saved first fields as it does so (see copyAndUnmark: and copyAndUnmarkMobileObjects). If the forwarding fields of live objects in the to-be-moved portion of the entire heap won't fit in savedFirstFieldsSpace, then additional passes can be made until the entire heap has been compacted.  When snapshotting multiple passes are made, but when doing a normal GC only one pass is made.

Each pass uses a three finger algorithm, a simple extension of the classic two finger algorithm with an extra finger used to identify the lowest pinned object between the to and from fingers.  Objects are moved down, starting at the first free object or chunk, provided that they fit below the lowest pinned object above the to finger.  When an object won't fit the to finger is moved above the pinned object and the third finger is reset to the next pinned object below the from finger, if any.

\subsection{Mark-Sweep}

Cog's Mark-Compact algorithm, named \emph{SpurSweeper}, is a sweep-only algorithm, setting the compactor to SpurSweeper effectively changes the fullGC to a mark-sweep non-moving algorithm.

SpurSweeper is a reference implementation if one wants to evaluate GC performance and compare it to a Mark-Sweep. It's also the only non-moving GC available right now which can be convenient for some experiments. One of the main reason why it was implemented is because advanced compaction algorithm includes a sweep phase (See SelectiveCompactor for example) and SpurSweeper allows to debug the sweep phase separately.

\subsection{Snapshot discussion}
Problem with the snapshot. and non compacting algo. Mention G1 and partial compaction too.

=> HybridCompactor

\section{Building Macro-benchmarks}
\label{sec:bench}

2 directions. One is to build real benchs from real apps. Same direction as Dacapo... To be ported ?

\subsection{Moose benchmark}

Explain all the phases

> Growing heap:
[GCMooseBench new benchLoadFromMSE: b wildfly].
benchLoadFromMSE: fileName
mooseModel := MooseModel new importFromMSEStream: (StandardFileStream readOnlyFileNamed: fileName)

[(GCMooseBench withMooseModel: model) benchExpandProperties]. "optional, increases drastically the model size"
benchExpandProperties
	allProperties :=  mooseModel entityStorage collect: [ :anEntity | 
    		anEntity mooseInterestingEntity complexPropertyPragmas collect: [ :aPragma |
        		[ (anEntity mooseInterestingEntity perform: aPragma methodSelector) ] on: Error do: ['error'] ] ].

> Stable heap:
[(GCMooseBench withMooseModel: model) benchAnalysisPyramid].
benchAnalysisPyramid
	mooseModel overviewPyramidMetrics.
[(GCMooseBench withMooseModel: model) benchAnalysisSymbol].
benchAnalysisSymbol
	mooseModel allMethods symbolsUsedInName
[(GCMooseBench withMooseModel: model) benchAnalysisDeprecation].
benchAnalysisDeprecation
	mooseModel allModelClasses select: [ :each |
    		(each isAnnotatedWith: 'Deprecated') and: [ each clientTypes notEmpty ] ].

> Shrinking heap:
[(GCMooseBench withMooseModel: model) benchRelease].
benchRelease
	mooseModel := nil.
	allProperties := nil.
	3 timesRepeat: [Smalltalk garbageCollect].

\subsection{Configurable benchmark}

Sophie's section

\section{Experimental results}
\label{sec:valid}

It would be nice to evaluate both Sweeper and Planning on what we've just discussed.

\section{Related and Future Work}

\paragraph{GC benchmark suite.}
Dacapo + see ref in Dacapo.

\paragraph{ACDC.}
ACDC-js (Sophie you write a paragraph on that.

\paragraph{Future work: more benchmarks.}
need to add more benchmarks based on Dacapo ?

\paragraph{Future work: stress GC tests.}
stress cases from Configurable
